---
title: "Why Multilevel Models Are Good Models For Longitudinal Data"
subtitle: "Multilevel Models Offer An Incredibly Flexible Treatment of Time and Time Varying Processes and Covariates"
author: "Andy Grogan-Kaylor"
date: "today"
format:
  html:
    toc: true
    number-sections: true
    anchor-sections: true
    theme:
      light: yeti
      dark: vapor
  pdf:
    toc: true
    number-sections: true
    geometry: margin=1in
    fig-height: 3
editor: source
bibliography: MLM.bib
---

# Visually

```{r}
#| echo: false

set.seed(3846)

N <- 10 # individuals

N_t <- 3 # time points

id <- rep(seq(1, N), each = N_t)

t <- runif(N * N_t, min = 1, max = 10)

u0 <- id

e <- rnorm(N * N_t, 0, 1)

y <- t + u0 + e

mydata <- data.frame(id, t, y)

mydata$y[1] <- NA # wave missing

mydata$y[11] <- NA # wave missing

```

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-MLM
#| fig-cap: "A Multilevel Model For Longitudinal Data"

library(ggplot2)

mycaption <- paste("Individuals Start And End The Study At Different Times", 
                   "\n",
                   "Every Individual Has Different Time Points",
                                      "\n",
                   "Individuals Have Different Numbers of Time Points")

ggplot(mydata,
       aes(x = t,
           y = y, 
           color = factor(id))) +
  geom_point(size = 3) +
  geom_smooth(method="lm",
              se = FALSE) +
  theme_minimal() +
  scale_color_viridis_d(name = "individual",
                        option = "turbo") +
  labs(title = "A Multilevel Model For Longitudinal Data",
       subtitle = "Every Individual Has Their Own Regression Line", 
       caption = mycaption)

```

# Data Structures

Multilevel models for longitudinal data prefer data in long format.

| id  | x1  | x2  | x3 | y1 | y2 | y3 |
|-----|-----|-----|----|----|----|----|
| 1   |     |     |    |    |    |    |
| 2   |     |     |    |    |    |    |
| 3   |     |     |    |    |    |    |

: Data in WIDE format {#tbl-wide}

| id  | t   | x   |  y  |
|-----|-----|-----|-----|
| 1   | 1   |     |     |
| 1   | 2   |     |     |
| 1   | 3   |     |     |
| 2   | 1   |     |     |
| 2   | 2   |     |     |
| 2   | 3   |     |     |
| 3   | 1   |     |     |
| 3   | 2   |     |     |
| 3   | 3   |     |     |

: Data in LONG format {#tbl-long}

# Equation

$$y_{it} = \beta_0 + \beta_1 t_{it} + \beta_2 x_{it} + u_{0i} + e_{it}$$ {#eq-longitudinal-mlm}

::: {.callout-tip}
## Person-Observations

Every row is a *person-observation* (person *i* observed at time *t*). Every person has *multiple rows*.
:::

# Advantages of the Multilevel Model for Longitudinal Data

1. Using the algebra in @eq-longitudinal-mlm, these models can easily accommodate both time varying and time invariant coefficients [@Hox2010; @Hox2018; @Singer2003; @Raudenbush2002; @RabeHesketh2022].
2. There is no multicollinearity issue with multiple $\beta$ coefficients for multiple waves of data. By inspection of @eq-longitudinal-mlm, we see that there is only a single $\beta$ coefficient for each variable, $\therefore$ no multicollinearity problem.
3. *Unbalanced data is less of a problem*, the data structure and estimation are robust to these possibilities [@Singer2003; @Raudenbush2002].
4. *Missing data is less of a problem *(assuming *MCAR*). When a person observation is missing, that person simply has fewer rows of data [@Hox2010; @Luke2004; @Raudenbush2002; @RabeHesketh2022]. But all rows of data are "matched" to the same person by $i$.
5.  We have an *explicit function of time* $\beta_1 t$, and could treat time more flexibly, by creating a polynomial function of time e.g. by adding $\beta_2 t^2$, etc. [@Raudenbush2002; @Singer2003]. (We could even substitute $\beta \ln(t)$.)
6.  Again, by inspection of @eq-longitudinal-mlm, we see that *multiple or many time-points are not a problem*. We would use the same algebra for 2 time points or for 10,000 time points. (Helpful when we start to think about intensive longitudinal data *e.g.* George Holden's *recording study*).
7.  We are *measuring exactly the time at which events take place* for each individual [@Singer2003; @Luke2004]. Not simply saying *Wave 1*, *Wave 2*, *Wave 3*, etc...
8.  Every individual could have a *completely different set of time points* and even a *completely different number of time points* [@Hox2010; @Hox2018; @Singer2003; @Luke2004].

::: {.callout-caution}
We do need to think carefully about what is the appropriate variable for time. Is it the variable we used to reshape the data--often `wave`--or some other more appropriate metric, like `age` or `time in study` [@Singer2003]?
:::
