---
title: "Multiple Methods of Longitudinal Data Analysis"
subtitle: "DRAFT Notes"
author: "Andy Grogan-Kaylor"
date: "today"
format:
  html:
    theme: 
      light: cosmo
      dark: vapor
    toc: true
    number-sections: true
    highlight-style: haddock
    code-fold: true
  pdf: 
    toc: true
    number-sections: true
    highlight-style: haddock
---

# Some Data

```{r}
#| output: false
#| echo: false

set.seed(3846)

library(Statamarkdown)

```

```{stata, collectcode = TRUE}

clear all

set seed 3846 //  set random seed

quietly set obs 10 // 10 observations

generate id = _n // id number

quietly expand 3 // expand by 3

sort id // sort by id

bysort id: generate t = _n // time variable

generate x = rnormal(10, 3) // random normal variable

generate w = rbinomial(1, .3) // random binomial variable

generate e = rnormal(0, 1) // random error

generate y = x + w + e // regression equation

drop e // drop error

list // list out the data

save longitudinal.dta, replace

```

# Multilevel Modeling

::: {.panel-tabset}

## Equation

$y_{it} = \beta_0 + \beta_1 x_{it} + \beta_2 w_{it} + u_{0i} + e_{it}$

::: column-margin
We assume that $u_{0i}$ has a normal distribution, but do not directly estimate the values of $u_{0i}$ for each individual.
:::

```{r}
#| column: margin
#| echo: false

hist(rnorm(1000, 0, 1),
     main = expression(u['0i']),
     xlab = "",
     col = "lightblue",
     yaxt="n",
     cex.main = 3.0)

```

## Syntax And Results

```{stata}

use longitudinal.dta, clear

mixed y x i.w || id:

```

:::

# Fixed Effects

::: {.panel-tabset}

## Equation

$y_{it} = \beta_0 + \beta_1 x_{it} + \beta_2 w_{it} + u_{0i} + e_{it}$

::: column-margin
We assume that the $u_{0i}$ are in fact, estimable. However, we end up estimating $y_{it} - \bar y_i = \beta_1 (x_{it} - \bar x_i) + \beta_2 (w_{it} - \bar w_i) + (e_{it} - \bar e_i)$. The $u_{0i}$ have dropped out of this equation.
:::

## Syntax And Results

```{stata}

use longitudinal.dta, clear

xtreg y x i.w, i(id) fe

```

:::

# Difference in Differences

???

# Cross Lagged Regression

::: {.panel-tabset}

## Equation

Similar to before, there is an equation predicting *y*.

$y_{2i} = \beta_0 + \beta_1 y_{1i} + \beta_2 x_{1i} + \beta_3 w_{2i} + e_i$

However, we need an equation for each time point, so:

$y_{3i} = \beta_0 + \beta_1 y_{2i} + \beta_2 x_{2i} + \beta_3 w_{2i} + e_i$

And, there are also equations predicting *x*.

$x_{2i} = \beta_0 + \beta_1 x_{1i} + \beta_2 y_{1i} + \beta_3 w_{1i} + e_i$

$x_{3i} = \beta_0 + \beta_1 x_{2i} + \beta_2 y_{2i} + \beta_3 w_{2i} + e_i$

In cross-lagged regression, we need the data to be in wide format rather than long format.

## Data Wrangling

```{stata}

use longitudinal.dta, clear

reshape wide y x w, i(id) j(t) // reshape data to wide

save longitudinalWIDE.dta, replace

```

## Syntax And Results

```{stata}

use longitudinalWIDE.dta, clear

sem (y2 <- y1 x1 w1) ///
(x2 <- x1 y1 w1) ///
(y3 <- y2 x2 w2) ///
(x3 <- x2 y2 w2)  

```

:::

# Summary [^1]

[^1]: Some of the decisions in this table are arguable.

\begingroup
\fontsize{5pt}{7pt}\selectfont

```{r}

Method <- c("Multilevel Modeling",
            "Fixed Effects",
            "Cross Lagged Regression")

`Control for Time Invariant Observed` <- c("yes",
                                           "yes",
                                           "yes")

`Control for Time Varying Observed` <- c("yes",
                                         "yes",
                                         "yes")

`Control for Time Invariant Unobserved` <- c("partially",
                                             "yes",
                                             "no")

`Control for Time Varying Unobserved` <- c("no",
                                           "no",
                                           "no")
`Estimate Reciprocal Causality` <- c("no",
                                     "no",
                                     "yes")

`Control for Earlier or Baseline y` <- c("automatic",
                                         "automatic",
                                         "must explicitly specify")

mytable <- data.frame(Method,
                      `Control for Time Invariant Observed`,
                      `Control for Time Varying Observed`,
                      `Control for Time Invariant Unobserved`,
                      `Control for Time Varying Unobserved`,
                      `Estimate Reciprocal Causality`,
                      `Control for Earlier or Baseline y`)

pander:: pander(mytable)

```

\endgroup










